---
title: "QBS 103 Final Project V1"
author: "Elizabeth Chin"
date: "2023-07-23"
output: pdf_document
---

## Gene: ABCA1

# This gene encodes a protein located in the cell membrane, and it belongs to a group of proteins called ATP-binding cassette (ABC) transporters. These proteins are responsible for moving various molecules in and out of cells. There are seven different subgroups of ABC genes, and this particular protein belongs to the ABC1 subgroup, which is unique to multicellular organisms. This protein plays a role in removing cholesterol from cells by acting as a pump to push cholesterol out of the cell. When both copies of this gene have mutations, it leads to conditions known as Tangier disease and familial high-density lipoprotein (HDL) deficiency, affecting how cholesterol is processed in the body.


```{r}
#set working directory
setwd("/Users/li_li/desktop/103_Final_Project/Data")

#read in csv files
gene.exp <- read.csv("QBS103_finalProject_geneExpression.csv", row.names = 1)

meta.data <- read.csv("QBS103_finalProject_metadata.csv", row.names = 1)
```

```{r}
#subset gene expression ABCA1
gene.ABCA1 <- gene.exp['ABCA1', ]

#transform rows to columns 
gene.ABCA1 <- as.data.frame(t(gene.ABCA1))
head(gene.ABCA1)
```

```{r}
#column bind two data frames
ABCA1.gene.exp.meta <- cbind(gene.ABCA1,meta.data)
head(ABCA1.gene.exp.meta, 1)
```

```{r}
#convert age to an integer
ABCA1.gene.exp.meta$age <- as.integer(ABCA1.gene.exp.meta$age) 
head(ABCA1.gene.exp.meta, 1)
```

```{r}
#creating histogram for gene expression

#load the required package
library(ggplot2)

#define color palette
my_palette1 <- c("#A2CFFE") # Choose any colors you like

#create theme
newBlankTheme <- theme(# Remove all the extra borders and grid lines
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        # Define my axis
        axis.line = element_line(colour = "black", linewidth = rel(1)),
        # Set plot background
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank(),
        legend.key = element_rect(fill = 'white'), 
        # Move legend
        legend.position = 'top')

#create histogram
ggplot(ABCA1.gene.exp.meta, aes(x=ABCA1.gene.exp.meta$ABCA1)) + 
  geom_histogram(binwidth = 2, fill = my_palette1, col="black") +
  labs(x = 'Gene Expression ABCA1',y = 'Count of ABCA1') +
  newBlankTheme 

```

```{r}
#create scatterplot for gene expression and age 

#define color palette
my_palette2 <- c("#ECCAFF") # Choose any colors you like

#create scatterplot
ggplot(ABCA1.gene.exp.meta,aes(x=ABCA1.gene.exp.meta$age,y=ABCA1.gene.exp.meta$ABCA1)) + 
  geom_point(color=my_palette2) + 
  labs(x = "Age", y = "Gene Expression ABCA1", title = "Gene Expression ABCA1 By Age") +
  newBlankTheme
```


```{r}
#create boxplot for gene expression separated by sex and icu status

#load the required package
library(dplyr)

#define color palette
my_palette3 <- c("#FFEEC4","#B1E3C7") # Choose any colors you like


#create a new data frame to filter the unknown in sex
filtered_sex <- ABCA1.gene.exp.meta %>% 
  filter(!grepl('unknown', sex)) 

#plot new data frame to create box plot
ggplot(filtered_sex, aes(x=filtered_sex$sex, y=filtered_sex$ABCA1, fill=filtered_sex$icu_status)) +
  geom_boxplot() +
  scale_fill_manual(values = my_palette3) +
  labs(x = "Sex",
       y = "Gene Expression ABCA1",
       fill = "ICU Status",
       title = "Gene Expression ABCA1 By Sex & ICU Status") + 
  newBlankTheme
  
```



